\documentclass{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{listings}
\usepackage{enumitem}

%--------------------------------------------------------------------
%  Mathematical specification of the quadratic extension field CM31
%  (a + b i with i^2 = -1) over GF(2^{31}-1)
%  Target language for implementation: Haskell
%--------------------------------------------------------------------

\begin{document}

\section{Base field recap}

Let
\[
   P = 2^{31}-1 = 2147483647
   \quad\text{and}\quad
   \mathbb{F}_{P} = \{\,0,\dots,P-1\,\}
\]
with canonical representation, reduction and arithmetic
as specified in the \texttt{M31} document.

\section{Quadratic extension}

\subsection{Irreducible polynomial}

Because \(P \equiv 3 \pmod 4\), the element \(-1\) is
\emph{not} a quadratic residue in \(\mathbb{F}_{P}\).  
Hence the polynomial
\[
   f(x)=x^{2}+1\in\mathbb{F}_{P}[x]
\]
is irreducible, and the quotient ring
\[
   \mathbb{K} := \mathbb{F}_{P}[x]/(x^{2}+1)
\]
is a field of order \(P^{2}\).

\subsection{Canonical representation}

Choose the integral basis \(\{1,i\}\) with \(i^{2}=-1\).
An element is written
\[
   z = a + b\,i,
   \quad a,b\in\mathbb{F}_{P},
\]
and represented concretely by the ordered pair
\[
   \operatorname{rep}(z) = (a,b)\in\mathcal{C}\times\mathcal{C},
   \qquad\mathcal{C}=[0,P-1].
\]
For serialisation use a little-endian byte layout

\[
   \text{bytes}(z)=\verb|LE32(a)|\;\Vert\;\verb|LE32(b)|\in\{0,\dots,255\}^{8}.
\]

\section{Core operations}

Throughout, \(a,b,c,d\in\mathcal{C}\).

\begin{enumerate}[label=\textbf{O\arabic*}.]
\item \textbf{Addition, negation, subtraction}
      \[
        (a,b) \oplus (c,d) = (a\!+\!c,\; b\!+\!d),
        \qquad
        \ominus(a,b) = (-a,\,-b),
        \qquad
        (a,b)\ominus(c,d)=(a\!-c,\;b\!-d),
      \]
      where \(\pm\) are the corresponding \texttt{M31} operations.

\item \textbf{Multiplication}
      \[
        (a,b)\otimes(c,d)
        \;=\;
        (\,a\,c - b\,d,\; a\,d + b\,c\,).
        \tag{$*$}
      \]
      Cost: \(4\) \texttt{M31} multiplies \(+\;2\) additions/subtractions.

\item \textbf{Squaring}
      \[
        (a,b)^{2} = (a^{2}-b^{2},\;2ab).
      \]

\item \textbf{Conjugation}
      \[
        \overline{(a,b)} = (a,-b).
      \]

\item \textbf{Norm}
      \[
        N(a,b) = (a,b)\,\overline{(a,b)}
                = (a^{2}+b^{2},\,0)\in\mathbb{F}_{P}\subset\mathbb{K}.
      \]

\item \textbf{Inverse}\;(\(z\neq0\))
      \[
        z^{-1} = \dfrac{\overline{z}}{N(z)}
               \;=\;
        \bigl(a,-b\bigr)\;*\;N(a,b)^{-1},
      \]
      where \(N(a,b)^{-1}\) is the \texttt{M31} inverse of \(a^{2}+b^{2}\).

\item \textbf{Embedding of \(\mathbb{F}_{P}\)}
      \[
         \iota\colon\mathbb{F}_{P}\hookrightarrow\mathbb{K},
         \qquad
         \iota(a)=(a,0).
      \]
\end{enumerate}

\section{Field axioms}

For all \(x,y,z\in\mathbb{K}\) the seven axioms (commutativity,
associativity, identities, inverses, distributivity)
listed for \texttt{M31} hold \emph{component-wise} together with
\[
   i^{2} = -1
   \quad\text{and}\quad
   \overline{\overline{z}}=z,\;
   z\,\overline{z}=N(z).
\]

\section{Complexities}

\[
\begin{array}{l|c}
\text{operation} & \text{M31 cost} \\ \hline
\oplus,\ominus & 2\text{ adds} \\
\otimes & 4 \text{ mults} + 2 \text{ adds} \\
\text{inverse} & 1\text{ conj} + 2\text{ sqrs} + 1\text{ add}
                 + \text{M31.inverse} + 2\text{ mults}
\end{array}
\]

\section{Testing specification (Haskell)}

\subsection{Property-based tests (QuickCheck)}

\begin{enumerate}[label=\textbf{Q\arabic*}.]
\item \textbf{Canonical range}  
      \(\forall(a,b).\;0\le a,b < P\).
\item \textbf{Field laws}  
      Verify axioms P1–P7 for \(100\,000\) random triples of \(\mathbb{K}\).
\item \textbf{Unit element}  
      \(\forall z.\; z\otimes 1 = z \land 1\otimes z = z\).
\item \textbf{Conjugation involution}  
      \(\forall z.\;\overline{\overline{z}}=z\).
\item \textbf{Norm property}  
      \(\forall z.\;z\otimes\overline{z} = N(z)\).
\item \textbf{Inverse correctness}  
      \(\forall z\neq0.\;z\otimes z^{-1}=1\).
\item \textbf{$i^{2}=-1$}  
      \((0,1)^{2}=(-1,0)\).
\item \textbf{Embedding consistency}  
      \(\forall a\in\mathbb{F}_{P}.\;\iota(a)\) behaves
      as a subfield element w.r.t.\ both \(\oplus\) and \(\otimes\).
\end{enumerate}

\subsection{Deterministic unit tests}

\begin{enumerate}[label=\textbf{T\arabic*}.]
\item Check the Rust examples verbatim  
      \( (1,2)^{-1}(1,2)= (1,0)\), etc.
\item Edge cases \((0,0),(1,0),(0,1),(P\!-\!1,P\!-\!1)\).
\item Serialisation: round-trip
      \(100\) random elements through the 8-byte
      little-endian format.
\item Attempt cast \((a,b)\to\mathbb{F}_{P}\) succeeds
      iff \(b=0\).
\end{enumerate}

\subsection{Performance goals}

\begin{itemize}[nosep]
\item Addition/negation \(<\!25\)\,ns.
\item Multiplication \(<\!120\)\,ns.
\item Inversion \(<\!2\)\,$\mu$s.
\end{itemize}

\section{Haskell implementation hints}

\begin{itemize}[nosep]
\item Re-use the previously defined
      \verb|newtype M31 = M31 Word32|.
\item Define
      \verb|data CM31 = CM31 {-# UNPACK #-}!M31 {-# UNPACK #-}!M31|.
\item Derive \verb|Show|, \verb|Eq|, and implement
      \verb|Num|, \verb|Fractional|, \verb|Enum|, \verb|Random|.
\item Inline small functions and mark
      \verb|CM31| as \verb|Storable|/ \verb|Prim| for SIMD-friendly code.
\item Use RULES to fuse \verb|conjugate . conjugate| into \verb|id|.
\item Consider Karatsuba if multiplication
      dominates the workload (rare for degree 2).
\end{itemize}

\end{document}
